---
title: "A Tutorial for select.surrogate: Treatment-Aware Regularization"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

`select.surrogate` estimates the proportion of treatment effect explained (PTE) by a set of candidate surrogate markers, and optionally first selects a subset of those candidates. It implements the treatment-aware regularization (TAR) approach of Parast, Freestone, and Tarr (2026), along with two related approaches for comparison: using all candidates with no selection, and an unweighted regularized approach.

Three selection options are available, set via the `selection` argument:

- `selection = "none"`: no selection; the PTE is estimated using all candidate markers, agreeing with `R.s.estimate` run with `number = "multiple"`.
- `selection = "TAR"`: treatment-aware regularization. Candidate markers with little or no treatment effect are penalized more heavily in a weighted lasso fit to the treated group, and the penalty is chosen, via cross-validation within the treated group, to be the sparsest point along the path whose estimated PTE falls no more than a pre-specified tolerance (`epsilon`) below that of the full candidate set.
- `selection = "rr"`: an unweighted, ordinary regularized approach (robust regularization), included mainly for comparison to TAR.

This tutorial walks through all three options using the example dataset included with the package, `d_tar_example`, and briefly covers the remaining optional arguments.

```{r setup}
#you will need version 3.3 or higher of the package
#if using CRAN version
#install.packages("RSurrogate")
#if using Github
#library(remotes
#remotes::install_github("laylaparast/RSurrogate")

library(Rsurrogate)

#set seed
set.seed(100)
```

## Example data

`d_tar_example` is one simulated dataset with 500 treated and 500 control observations, and 9 candidate surrogate markers. By design, only the first three markers (`s1`, `s2`, `s3`) both differ between the treatment groups and predict the outcome; the remaining six markers are not part of the true minimal surrogate set.

```{r data}
data(d_tar_example)
names(d_tar_example)
dim(d_tar_example$sone)
dim(d_tar_example$szero)
length(d_tar_example$yone)
length(d_tar_example$yzero)
```

`sone` and `szero` are matrices of the 9 candidate markers for the treated and control groups, respectively, and `yone` and `yzero` are the corresponding primary outcome vectors. This is exactly the argument structure used by `R.s.estimate` elsewhere in this package.

## No selection: using all candidate markers

Setting `selection = "none"` estimates the PTE using all 9 candidate markers, with no selection step.

```{r none}
fit.none <- select.surrogate(sone = d_tar_example$sone, szero = d_tar_example$szero,
                              yone = d_tar_example$yone, yzero = d_tar_example$yzero,
                              selection = "none")
fit.none
```

`delta` is the estimated overall treatment effect, `delta.s` is the estimated residual treatment effect after accounting for all 9 markers, and `R.s` is the resulting PTE estimate. There is no `selection` element here, since no markers were removed.

## Treatment-aware regularization (TAR)

Setting `selection = "TAR"` instead first selects a subset of the 9 candidate markers.

```{r tar}
fit.tar <- select.surrogate(sone = d_tar_example$sone, szero = d_tar_example$szero,
                             yone = d_tar_example$yone, yzero = d_tar_example$yzero,
                             selection = "TAR")
fit.tar
names(fit.tar$selection)
```

`fit.tar$selection` is a named vector giving the (penalized) lasso coefficient of each selected marker; `names(fit.tar$selection)` gives just the names of the markers TAR selected. In this example, TAR tends to recover `s1`, `s2`, and `s3`, the three markers that actually drive the treatment effect, while `R.s` remains close to the full-candidate-set estimate from `fit.none` even though far fewer markers are used.

## An unweighted regularized approach (rr)

Setting `selection = "rr"` uses an ordinary (unweighted) cross-validated lasso instead, with no treatment-effect-based penalty weighting.

```{r rr}
fit.rr <- select.surrogate(sone = d_tar_example$sone, szero = d_tar_example$szero,
                            yone = d_tar_example$yone, yzero = d_tar_example$yzero,
                            selection = "rr")
fit.rr
names(fit.rr$selection)
```

Because `rr` selects markers based only on how well they predict the outcome, without regard to whether they differ between treatment groups, it will often select a different, sometimes larger, set of markers than TAR, and can achieve a lower PTE for a given set size.

## Comparing the three approaches

```{r compare}
data.frame(
  approach   = c("none", "TAR", "rr"),
  n.selected = c(ncol(d_tar_example$sone), length(fit.tar$selection), length(fit.rr$selection)),
  R.s        = c(fit.none$R.s, fit.tar$R.s, fit.rr$R.s)
)
```

## Other options

### Bootstrap variance and confidence intervals

Setting `var = TRUE` adds a bootstrap standard error and a 95% confidence interval (the 2.5th and 97.5th percentiles of the bootstrap replicates) for `delta`, `delta.s`, and `R.s`. Unlike the rest of this package, which uses perturbation-resampling, `select.surrogate` uses a nonparametric bootstrap: both treatment groups are resampled with replacement and the entire selection procedure, including re-selection of the penalty, is rerun on every resample. `boot.num` controls the number of resamples (200 by default, matching the simulation study in Parast, Freestone, and Tarr (2026)). Because the whole procedure is rerun on every resample, this can take a while, especially for `selection = "TAR"`; a smaller `boot.num` is used below just for a quick illustration.

```{r var}
fit.tar.var <- select.surrogate(sone = d_tar_example$sone, szero = d_tar_example$szero,
                                 yone = d_tar_example$yone, yzero = d_tar_example$yzero,
                                 selection = "TAR", var = TRUE, boot.num = 50)
fit.tar.var$R.s.var
fit.tar.var$conf.int.R.s
```

### The lasso path: nlambda

`nlambda` (default 20) controls the number of penalty values considered along the lasso path for `selection = "TAR"` or `selection = "rr"`. A finer grid (larger `nlambda`) gives TAR or rr more candidate penalty values to choose from, at the cost of more computation.

### Tuning TAR: folds, epsilon, c.constant, small.n

These four arguments only affect `selection = "TAR"`:

- `folds` (default 5): the number of cross-validation folds, within the treated group, used to select the penalty.
- `epsilon` (default 0.05): the tolerance. TAR selects the sparsest point along the weighted lasso path whose estimated PTE falls no more than `epsilon` below that of the full candidate set. A larger `epsilon` allows a smaller selected set at the cost of a larger potential drop in PTE.
- `c.constant` (default 1): controls how strongly a candidate marker's estimated treatment effect magnitude affects its penalty weight; larger values penalize low-treatment-effect markers more aggressively.
- `small.n` (default 1e-6): a small constant added before inverting each marker's estimated treatment effect magnitude to form its penalty weight, so that a marker with an estimated treatment effect of (or very close to) zero does not produce an infinite weight.

### Tuning rr: lambda_rule

`lambda_rule` (default `"lambda.1se"`) only affects `selection = "rr"`, and controls which cross-validated penalty is used: `"lambda.1se"`, the most regularized penalty within one standard error of the minimum cross-validated error, or `"lambda.min"`, the penalty that minimizes cross-validated error directly. `"lambda.1se"` is more conservative and more likely to select a smaller (sometimes empty) set of markers.

## Reference

Parast, L., Freestone, G., Tarr, J. (2026). Treatment-Aware Regularization via Multi-Objective Optimization for Surrogate Selection.
